\documentclass[a4paper]{exam}
%%% BEGIN PREAMBLE

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows,automata}
%\usepackage{vaucanson-g}
% \usepackage{beamerthemesplit} // Activate for custom appearance
\usepackage[german, english]{babel}
\usepackage[utf8]{inputenc}

% Definitions
\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{example}{Example}

\newcommand{\emptystring}{$\varepsilon$}
\newcommand{\length}[1]{\mid #1 \mid}
\newcommand{\directlygenerates}{\Rightarrow}
\newcommand{\generates}{\Rightarrow^{+}}
\newcommand{\generatesequal}{\Rightarrow^{*}}

\newenvironment{grammar}
	{\begin{tabular}[b]{lcl}}
	{\end{tabular}}
\newcommand{\rewritten}{$\to$}
\newcommand{\alternative}{$\mid \;$}

% Title stuff
\title{Regular Languages --- Exercises}
\author{Peter Bauer}
\date{Version\_01.01.02} % delete this line to display the current date

%%% BEGIN DOCUMENT
\begin{document}

%\printanswers

\begin{center} \fbox{\fbox{\parbox{5.5in}{\centering
IF.04.22 LOAL --- Exercises --- Regular Languages and DFA}}}
\end{center}
%\makebox[\textwidth]{Name: \enspace\hrulefill}

\begin{questions}
	\section{Interpreting Regular Expressions}
\question
	\textbf{Interpreting a Regular Expression}
	
	Find the shortest string $\omega \in \{a, b\}^*$ that is not in the language represented by the regular expression $a^*(ab)^*b^*$. Give
	reasons for your answer.
	
\question
	\textbf{Interpreting a Regular Expression}
	Consider the regular expression $(a(cd)^*b)^*$.
	\begin{parts}
		\part Find a string of length 2 over $\{a, b, c, d\}$ which matches the expression.
		\part Find a string of length 1 over $\{a, b, c, d\}$ which does not match the expression.
	\end{parts}
	
	\begin{otherlanguage}{german}
		Finden Sie für den oben angegebenen regulären Ausdruck
		\begin{parts}
			\part einen String der Länge 2 mit den Symbolen $\{a, b, c, d\}$, welcher als gültig erkannt wird
			\part einen String der Länge 1 mit den Symbolen $\{a, b, c, d\}$, welcher als ungültig erkannt wird
		\end{parts}
	\end{otherlanguage}
\question
	\textbf{Interpreting a Regular Expression}
	
	Describe in an English sentence (as simple as possible) the language corresponding to the regular expression:
	
	$(b + ab)^*(a + ab)^*$.

\begin{otherlanguage}{german}
	Beschreiben Sie in einem möglichst einfachen deutschen Satz die Sprache des oben gegebenen regulären Ausdrucks.
\end{otherlanguage}

\begin{solution}
A sequence of $b$s preceded or interrupted by single $a$s followed by a sequence of $a$s interrupted by single $b$s or empty.
\end{solution}

\section{C, C\#, Java Comments}
\question
	\textbf{Comments}
		In C, C\#, or Java comments can be given in the form {\tt /* This is the comment */}. This means that comments start
		with the string {\tt /*} and end with the string {\tt */}. The comment itself can be an arbitrary string with the exception that
		the string {\tt */} must not occur.
		
		More precisely we could say that the language of a comment is now the string starting with ``/'' followed by a ``*'' followed by an arbitrary long sequence of characters except the sequence ``*/'' and ending with the sequence ``*/''. You may write $\alpha - \{$``x''$\}$ to denote a set of characters excluding one specific character ``x''.
		
\begin{parts}
	\item Give a state diagram of a DFA which can parse this language.
	\item Give a regular grammar which describes the language of C comments.
	\item Describe this language by a regular set.
	\item Is it possible to describe nested comments, e.g., {\tt /* This is a /* nested */ comment */}, by a regular expression? Give arguments for your decision.
\end{parts}
		
\begin{otherlanguage}{german}
In C, C\# oder Java gibt es die Blockkommentare, welche in \verb$/*$ und \verb$*/$ eingeschlossen sind. Der Kommentar selbst kann eine beliebige Zeichenkette sein, welche nur dadurch eingeschränkt ist, dass sie die Sequenz \verb$*/$ nicht beinhalten darf.

\begin{parts}
	\item Geben Sie das Zustandsdiagramm eines DFA an, der diese Sprache erkennt.
	\item Geben Sie die dazugehörige reguläre Grammatik an.
	\item Geben Sie den dazugehörigen regulären Ausdruck an.
	\item Ist es möglich, geschachtelte Kommentare in Form eines regulären Ausdrucks zu beschreiben? Begründen Sie Ihre Antwort.
\end{parts}

\end{otherlanguage}

\begin{solution}
	We denote the character class of all valid characters for a program code by $a$. Let's define the following character classes:
	
	\begin{align}
		s &= a - \{"/"\} \\
		e & = a - \{"*"\} \\
		c &= a - \{"*", "/"\} \\
		r &= a - \{"/"\}
	\end{align}

	Then the grammar looks like this:

	\begin{grammar}
		S & \rewritten & sS \alternative /A \\
		A & \rewritten & eS \alternative *C \\
		C & \rewritten & cC \alternative *U \\
		U & \rewritten & rC \alternative / \alternative /F \\
		F & \rewritten & aF
	\end{grammar}
	
	RegEx: {\tt /\textbackslash*a\textbackslash*/}
\end{solution}

\section{$a$s and $b$s sometimes $c$s}
\question
	\textbf{\texttt{aab}s and \texttt{bb}s}
	
	Given a language $L$ defined over the alphabet $\{\mathtt{a}, \mathtt{b}\}$ defined recursively as follows:
	\begin{enumerate}
		\item $\varepsilon \in L$.
		\item If $x \in L$ then $\mathtt{aab}x \in L$ and $x\mathtt{bb} \in L$.
		\item Nothing else is in $L$.
	\end{enumerate}
	
	Please answer the following questions concerning $L$:
	\begin{parts}
		\item Draw the state diagram of the DFA parsing strings of this language.
		\begin{solution}
			We start by denoting a few first strings, which are part of the language.
			
			\begin{tabular}{cccccccc}
				\multicolumn{8}{c}{$\varepsilon$} \\
				\multicolumn{4}{c}{$aab$} & \multicolumn{4}{c}{$bb$} \\
				\multicolumn{2}{c}{$aabaab$} & \multicolumn{2}{c}{$aabbb$} & \multicolumn{2}{c}{$aabbb$} & \multicolumn{2}{c}{$bbbb$} \\
				$aabaabaab$ & $aabaabbb$ & $aabaabbb$ & $aabbbbb$ & $aabaabbb$ & $aabbbbb$ & $aabbbbb$ $bbbbbb$ 
			\end{tabular}
			
			We see, that the strings could start with a sequence of $aab$s which can be repeated but as soon as we enter in a $bb$, it never goes back to a $aab$.
		\end{solution}
		\item Give the regular grammar for $L$.
		\item Give the regular expression which describes $L$.
	\end{parts}

	\question
	\textbf{$4i + 1 b$'s}
	Given the following regular language:
	
	\begin{tabular}[b]{lcl}
		$L(G(S))$ & = & $\{\omega \in (a \mid b \mid c)^* \mid \omega$ has $4i + 1 b$'s ($i \geq 0$)\}
	\end{tabular}

	\begin{parts}
		\item Give the state diagram of a DFA which parses this language.
		\item Give a regular grammar $G(S)$ which describes this language.
		\item Give a regular expression which describes this language.		
	\end{parts}
	
	\begin{solution}
	
	\begin{grammar}
		S & \rewritten &  \alternative aS \alternative b \alternative bA \alternative cS \\
		A & \rewritten & a \alternative c \alternative aA \alternative cA \alternative bB \\
		B & \rewritten & aB \alternative cB \alternative bC \\
		C & \rewritten & aC \alternative cC \alternative bD \\
		D & \rewritten & aD \alternative cD \alternative b \alternative bF \\
		F & \rewritten & a \alternative c \alternative aF \alternative cF \alternative bB
	\end{grammar}
	
	RegEx: {\tt (a|c)*b((a|c)*(bbbb)*(a|c)*)*}
	
	\end{solution}

\question \textbf{No $bbb$}
	Given the following regular language:
	
	\begin{tabular}[b]{lcl}
		$L(G(S))$ & = & $\{\omega \in (a \mid b)^* \mid \omega$ does {\em not} contain the substring $bbb\}$
	\end{tabular}

	\begin{parts}
		\item Give the state diagram of a DFA which parses this language.

		\item Give a regular grammar $G(S)$ which describes this language.
		
	\end{parts}
	
	\begin{otherlanguage}{german}
		Gegeben sei oben genannte reguläre Sprache.
		
		\begin{parts}
		\item Zeichnen Sie das Zustandsdiagramm eines DFA, der diese Sprache erkennt.

		\item Geben Sie die reguläre Grammatik $G(S)$ für diese Sprache an.
		
	\end{parts}

	\end{otherlanguage}

	\begin{solution}
		
		\begin{grammar}
			S & \rewritten & $\varepsilon$ \alternative a \alternative aF \alternative b \alternative bD \\
			F & \rewritten & a \alternative aF \alternative b \alternative bD \\
			D & \rewritten & a \alternative aF \alternative b \alternative bR \\
			R & \rewritten & a \alternative aF \alternative bE \\
			E & \rewritten & aE \alternative bE
		\end{grammar}
		
	\end{solution}

	\section{With 0s and 1s}
\question
	\textbf{Starts with two 0s}

	Given the following regular language:
	
	\begin{tabular}[b]{lcl}
		$L(G(S))$ & = & $\{\omega \in (0 \mid 1)^* \mid \omega$ starts with two 0s\}
	\end{tabular}

	\begin{parts}
		\item Give the state diagram of a DFA which parses this language.
		\item Give a regular grammar $G(S)$ which describes this language.
		\item Give a regular expression which describes this language.
	\end{parts}

	\begin{solution}
		
		\begin{grammar}
			S &\rewritten & 0A \alternative 1E \\
			A &\rewritten & 0 \alternative 1 \alternative 0F \alternative 1F \\
			F &\rewritten & 0 \alternative 1 \alternative 0F \alternative 1F \\
			E &\rewritten & 0E \alternative 1E
		\end{grammar}
		
		RegEx: {\tt 00(0|1)*}
		
	\end{solution}
	
\question
	\textbf{Ends with 00 or 01}

	Given the following regular language:
	\begin{tabular}[b]{lcl}
		$L(G(S))$ & = & $\{\omega \in (0 \mid 1)^* \mid \omega$ ends with 00 or 01\}
	\end{tabular}

	\begin{parts}
		\item Give the state diagram of a DFA which parses this language.
		\item Give a regular grammar $G(S)$ which describes this language.
		\item Give a regular expression which describes this language.
	\end{parts}
	
	\begin{otherlanguage}{german}
		Geben sei obige reguläre Sprache $L(G(S))$.
		
		\begin{parts}
			\item Zeichnen Sie das Zustandsdiagramm eines DFA, der diese Sprache erkennt.
			\item Geben Sie die reguläre Grammatik $G(S)$ für diese Sprache an.
			\item Geben Sie den regulären Ausdruck für diese Sprache an.
		\end{parts}
	\end{otherlanguage}

	\begin{solution}
	\begin{parts}
		\part See grammar below
		\part
		
		\begin{grammar}
		S & \rewritten & 1S \alternative 0P \\
		P & \rewritten & 0 \alternative 0Z \alternative 1 \alternative 1E\\
		Z & \rewritten & 0 \alternative 0Z \alternative 1 \alternative 1E \\
		E & \rewritten & 0P \alternative 1S\\
		\end{grammar}
		
		\part $(0 \mid 1)^*(01 \mid 00)$
	\end{parts}
	\end{solution}	
\end{questions}
\end{document}